Moisture effect on leaf litter alters non-additive flammability in mixed conifer forest

### 2019-12-17 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
WIP

Species differ in their litter flammability \cite{Fonda+Belanger+etal-1998,
  Fonda-2001, Scarff+Westoby-2006, Kane+Varner+etal-2008,
  Cornwell+Elvira+etal-2015}. Litter packing, measured as litter-bed density or
packing ratio, is an important control ignition and spread in models of surface
fire spread \citep{Weber-1991, Balbi+Santoni+etal-1999, Morvan+Larini-2001}.
Multiple studies have demonstrated that litter bed density controls the
flammability of dry litter fuels in seasonally dry climates where higher litter
density reduces flame spread rate. Across these studies, there is increasing
consensus that larger less densely packed litter beds increase oxygen
availability and burn with faster spread rates \citep{Scarff+Westoby-2006,
  Schwilk+Caprio-2011, Magalhaes+Schwilk-2012, Cornwell+Elvira+etal-2015,
  Varner+Kane+etal-2015, Kreye+Varner+etal-2018}. The litter packing is driven
by litter particle size and shape which is in turn influenced by leaf size, how
leaves or branches are abscised, and subsequent decomposition
\citep{Fonda+Belanger+etal-1998, Stephens+Finney+etal-2004,
  Scarff+Westoby-2006, Kane+Varner+etal-2008, Cornwell+Elvira+etal-2015,
  Zhao+Cornwell+etal-2016, Grootemaat+Wright+etal-2017,
  Prior+Murphy+etal-2018, Zhao+vanLogtestijn+etal-2019}.

These species-specific effects are increasingly well established. However, how
moisture dynamics modulate species specific flammability is poorly known and
could be driven by by leaf traits, emergent litter traits, or by both. Species
specific effects on litter flammability are recorded for the forests of the
southeastern US \citep{Nowacki+Abrams-2008} where species-specific litter
drying rates play a role in flammability feedbacks
\citep{Kreye+Varner+etal-2013}. Fire behavior consequences of individual
species traits have been recorded for western US forests as well: in mixed
conifer forest in California, where the fire regime is characterized by surface
fires fueled by plant litter, overstory tree composition drives variation in
local fire behavior \citep{Schwilk+Caprio-2011} as a result of differential
litter flammability across tree species and as a result of synergistic
interactions in multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}.
These flammability feedbacks may help explain historical fire frequency
patterns \citep{Schwilk+Caprio-2011}, but current estimates of litter
flammability in mixed conifer forests are based largely on burning dry litter.

Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuel-beds
\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however. Drying rates following snowmelt
or following rainfall during the fire season will influence the length of the
fire season, flammability over time, and the probability of sustained ignition,
higher spread rates and fuel consumption when ignition occurs. That species can
differ in litter drying rates has been demonstrated for at least one forest
system \citep{Kreye+Varner+etal-2013}. Such differential moisture content
across time may contribute to the often high spatial heterogeneity of surface
fires \citep{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.

Fuel moisture content through time is controlled by two processes, absorption
and retention \citep{Kreye+Kobziar+etal-2013}. Absorptive capacity represents
the potential maximum fuel moisture and is evaluated by the moisture content of
litter when saturated. Differences in moisture content over time since
saturation can indicate different abilities to retain moisture. Litter particle
geometry may have direct effects on maximum moisture absorption through
area-mass relationships and we also expect effects of leaf morphology,
especially leaf cuticle thickness \citep{Van_Wagner-1969}. Like absorption,
moisture retention should also be influenced by leaf and litter bed traits:
surface-area-to-volume ratio and litter bulk density will influence air flow
through the litter bed. Litter packing therefore effects fire behavior in two
ways: directly by its effect on oxygen availability for combustion, and
indirectly by its effect on dry-down rate. Both these pathways are affected by
leaf particle traits, especially leaf size \citep{Scarff+Westoby-2006}.
Fuel-beds composed of larger leaves with lower moisture retainability should
favor drying, thus increasing the probability of ignition and fire spread.

@@

[TODO: reseearch question here]

We investigated species specific litter moisture dynamics across eight tree
species in a temperate fire-prone forest in California. We determined the
maximum water retention and rate of moisture loss and examined how these vary
in relation to two leaf and litter bed traits [CHECK AND FIX]. We then examined
the influence of moisture on the flammability across these species.
Experimental burn trials of leaf litter at various moisture levels determined
how time since wetting influenced two axes of flammability characterized by
flame spread rate and total fuel consumed. Previous work described the
flammability of these species in monocultures and in mixtures for oven-dry
litter \citep{Magalhaes+Schwilk-2012} and found consistent positive
nonadditivity of flammability where fire behavior of mixtures was driven by the
most flammable species. Here, we examined if this positive nonadditivity was
maintained in mixtures of varying levels of fuel moisture. The burn trials also
performed with multi-species litter mixtures to evaluate if there are
nonadditive effects as observed in oven-dried litter. We tested if the
nonadditive effects on reported in previous studies occur under [TODO CHECK]
higher moisture levels, and whether fuel moisture effects
strengthen or moderate such nonadditivity.

@@

We produced dry-down curves to determine maximum water retainability and the
rate of moisture content loss for each species and for four distinct
three-species mixtures. These curves allowed us to determine the rate at which
a species loses moisture, approximated as exponential decay, and to determine
maximum water content following saturation and gravity draining. Samples from
the eight species (six replicates per species) and four mixtures (5 replicates
per mixture type) were placed in baskets approximately 45 x 45\,cm by 15\,cm
tall constructed of aluminum mesh with 0.52\,cm square openings. Samples were
450\,g dry weight litter per trial and litter depth was standardized to 10\,cm
by changing the horizontal dimensions of the basket. Weight was measured by
using a balance sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus,
OH). Samples were immersed in plastic storage bins filled with water for 24
hours to fully saturate them. After saturation, the baskets were removed from
the water, allowed to drain for 3 minutes, and samples weighed to measure
maximum moisture content. The samples were then allowed to dry in a controlled
environment kept at $21^\circ$C temperature and 30\% relative humidity,
assessed via an iButton sensor (Maxim Integrated) placed over the samples.
Litter baskets were arranged in a completely randomized design. During the
dry-down process three subsamples of 5--10\,g were taken at 24 hour intervals
and weighed. These subsamples were then oven dried for 24 hours at $100^\circ$C
and re-weighed to assess moisture content. This was repeated until the final
sample from a basket was less than 10\% moisture.

@@

The genera in this study displayed different patterns of both moisture
absorption and desorption, associated with different litter traits. In this
sample of species, those that have a greater absorptive capacity are also among
the most flammable in terms of spread rate when dry. This highlights the
importance of fuel moisture in modulating flammability. One shortcoming of past
work in this system \citep{Magalhaes+Schwilk-2012} was that flammability trials
tested only dry fuels. We therefore expected that rankings of species
flammability might change under higher moisture levels. For
example, one presumption of ours prior to this experiment was that the high
flammability previously reported for \emph{Q. kelloggii}
\citep{Magalhaes+Schwilk-2012} would disappear under higher fuel moisture
contents. This prediction was partly supported: \emph{Q. kelloggii} fails to
ignite until after 72 hours of drying in this study (Figs.
\ref{fig:spread_time} and \ref{fig:consume_time}). However, high moisture
content in this oak was unable to suppress ignition of mixtures to which it
contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
\ref{fig:mixture_obs_pred_consume}).

The dry down curves we measured demonstrate that after a rain event, the litter
of different species will exhibit different moisture contents over time as a
consequence of both differential initial absorption and of differences in
drying rate. This has consequences for flammability as a function of species
and time since wetting. Our flammability trials demonstrate that both flame
spread rate and fuel consumption increased with time since wetting but that the
rate of increase varied across these taxa. Only the pines are able to ignite
and support flame spread immediately after wetting (Fig.
\ref{fig:spread_time}). In contrast, \emph{Quercus} exhibited a threshold
response to drying with no ignition prior to 72 hours of drying but once
ignition is possible, having faster spread rates than the other taxa. The
Cupressaceae, which in our two species have dense litter but low maximum water
absorption had low flammability across moisture levels. This result is
consistent with the low flammability found in an Australian Cupressaceae that
have small litter particles and dense packing citep{Scarff+Westoby-2006,
  Bowman+Haverkamp-2018}. However, the densely packed litter is not a
consistent trait across the family because \citet{Cornwell+Elvira+etal-2015}
found relatively high litter flammability and low litter density for some Asian
Cupressaceae.

>>>>

Species differ in their litter flammability \cite{Fonda+Belanger+etal-1998,
  Fonda-2001, Scarff+Westoby-2006, Kane+Varner+etal-2008,
  Cornwell+Elvira+etal-2015} and fire regime can vary of small spatial scales in response to community composition and species-specific flammability \citep{Stephens-2001, Hille+Stephens-2005, Schwilk+Caprio-2011}. Litter packing, measured as litter-bed density or
packing ratio, is an important control ignition and spread in models of surface
fire spread \citep{Weber-1991, Balbi+Santoni+etal-1999, Morvan+Larini-2001}.
Multiple studies have demonstrated that litter bed density controls the
flammability of dry litter fuels in seasonally dry climates where higher litter
density reduces flame spread rate. Across these studies, there is increasing
consensus that larger less densely packed litter beds increase oxygen
availability and burn with faster spread rates \citep{Scarff+Westoby-2006,
  Schwilk+Caprio-2011, Magalhaes+Schwilk-2012, Cornwell+Elvira+etal-2015,
  Varner+Kane+etal-2015, Kreye+Varner+etal-2018}. The litter packing is driven
by litter particle size and shape which is in turn influenced by leaf size, how
leaves or branches are abscised, and subsequent decomposition
\citep{Fonda+Belanger+etal-1998, Stephens+Finney+etal-2004,
  Scarff+Westoby-2006, Kane+Varner+etal-2008, Cornwell+Elvira+etal-2015,
  Zhao+Cornwell+etal-2016, Grootemaat+Wright+etal-2017,
  Prior+Murphy+etal-2018, Zhao+vanLogtestijn+etal-2019}.

These species-specific effects are increasingly well established. However, how
moisture dynamics modulate species specific flammability is poorly known. Species
specific effects on litter flammability are recorded for the forests of the
southeastern US \citep{Nowacki+Abrams-2008} where species-specific litter
drying rates play a role in flammability feedbacks
\citep{Kreye+Varner+etal-2013}. Fire behavior consequences of individual
species traits have been recorded for western US forests as well: in mixed
conifer forest in California, where the fire regime is characterized by surface
fires fueled by plant litter, overstory tree composition drives variation in
local fire behavior \citep{Schwilk+Caprio-2011} as a result of differential
litter flammability across tree species and as a result of synergistic
interactions in multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}.
These flammability feedbacks may help explain historical fire frequency
patterns \citep{Schwilk+Caprio-2011}, but current estimates of litter
flammability in mixed conifer forests are based largely on burning dry litter.

Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuel-beds
\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however. Drying rates following snowmelt
or following rainfall during the fire season will influence the length of the
fire season, flammability over time, and the probability of sustained ignition,
higher spread rates and fuel consumption when ignition occurs. That species can
differ in litter drying rates has been demonstrated for at least one forest
system \citep{Kreye+Varner+etal-2013}. Such differential moisture content
across time may contribute to the often high spatial heterogeneity of surface
fires \citep{Hille+Stephens-2005, Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and
to variation in fire regime across landscapes \citep{Stephens-2001}.

Fuel moisture content through time is controlled by two processes, wetting (including absorption and adsorption) and drying rate (desorbtion) \citep{Anderson-1990, Kreye+Kobziar+etal-2013}. Absorptive capacity represents
the potential maximum fuel moisture and is evaluated by the moisture content of
litter when saturated. Differences in moisture content over time since
saturation (drying rate) can indicate different abilities to retain moisture. Litter particle
geometry may have direct effects on maximum moisture absorption through
area-mass relationships and we also expect effects of leaf morphology,
especially leaf cuticle thickness \citep{Van_Wagner-1969}. Like absorption,
moisture retention should also be influenced by leaf and litter bed traits:
surface-area-to-volume ratio and litter bulk density will influence air flow
through the litter bed. Litter packing therefore effects fire behavior in two
ways: directly by its effect on oxygen availability for combustion, and
indirectly by its effect on dry-down rate. Both these pathways are affected by
leaf particle traits, especially leaf size \citep{Scarff+Westoby-2006}.
Fuel-beds composed of larger leaves or leaves with lower moisture retainability should
favor drying, thus increasing the probability of ignition and fire spread.

@@

We asked if differences in litter moisture absorption and drying rates could
change the relative flammability of litter across species compared with
estimates based on dry fuels. We investigated these litter moisture dynamics
across eight tree species in a temperate fire-prone forest in California. We
determined the maximum water retention and rate of moisture loss and examined
how these vary in relation to specific leaf area and litter packing density.
Experimental burn trials of leaf litter at various moisture levels determined
how time since wetting influenced two axes of flammability characterized by
flame spread rate and total fuel consumed. Previous work described the
flammability of these species in monocultures and in mixtures for oven-dry
litter \citep{Magalhaes+Schwilk-2012} and found consistent positive
nonadditivity of flammability where fire behavior of mixtures was driven by the
most flammable species. We tested if the nonadditive effects on reported in
previous studies occur under higher moisture levels, and whether fuel moisture
effects strengthen or moderate any nonadditivity.

@@

We produced dry-down curves to determine maximum water retainability and the
rate of moisture content loss for each species and for four distinct
three-species mixtures. These curves allowed us to determine the rate at which
a species loses moisture, approximated as exponential decay, and to determine
maximum water content following saturation and gravity draining. Samples from
the eight species (six replicates per species) and four mixtures (5 replicates
per mixture type) were placed in baskets approximately 45 x 45\,cm by 15\,cm
tall constructed of aluminum mesh with 0.52\,cm square openings. Samples were
450\,g dry weight litter per trial and litter depth was standardized to 10\,cm
by changing the horizontal dimensions of the basket. Weight was measured by
using a balance sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus,
OH). Samples were immersed in plastic storage bins filled with water for 24
hours to fully saturate them. After saturation, the baskets were removed from
the water, allowed to drain for 3 minutes, and samples weighed to measure
maximum moisture content. The samples were then allowed to dry in a controlled
environment kept at $21^\circ$C temperature and 30\% relative humidity,
assessed via an iButton sensor (Maxim Integrated) placed over the samples.
Litter baskets were arranged in a completely randomized design. During the
dry-down process three subsamples of 5--10\,g were taken at 24 hour intervals
and weighed. These subsamples were then oven dried for 24 hours at $100^\circ$C
and re-weighed to assess moisture content. This was repeated until the final
sample from a basket was less than 10\% moisture. This experimental setup was designed to accelerate drying compared with field conditions because the baskets were suspended in dry air 2.8 -- 5.3$^\circ$C above the warmest month average temperatures for the closest weather stations to the collection sites (Grant Grove and Lodegpole stations, NOAA 1981-2010 normals, https://www.ncdc.noaa.gov/cdo-web/datatools/normals).

@@

The genera in this study displayed different patterns of both moisture
absorption and desorption, associated with different litter traits. In this
sample of species, those that have a greater absorptive capacity are also among
the most flammable in terms of spread rate when dry. Traits such as large leaves and high SLA may affect both absorptive capacity and flammability with a consequence that flammability rankings across species can change with time since wetting. One shortcoming of past
work in this system \citep{Magalhaes+Schwilk-2012} was that flammability trials
tested only dry fuels. We therefore expected that rankings of species
flammability might change under higher moisture levels. For
example, one presumption of ours prior to this experiment was that the high
flammability previously reported for \emph{Q. kelloggii}
\citep{Magalhaes+Schwilk-2012} would disappear under higher fuel moisture
contents. This prediction was partly supported: \emph{Q. kelloggii} fails to
ignite until after 72 hours of drying in this study (Figs.
\ref{fig:spread_time} and \ref{fig:consume_time}). However, high moisture
content in this oak was unable to suppress ignition of mixtures to which it
contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
\ref{fig:mixture_obs_pred_consume}).

The dry down curves we measured demonstrate that after a rain event, the litter
of different species will exhibit different moisture contents over time as a
consequence of both differential initial absorption and of differences in
drying rate. This has consequences for flammability as a function of species
and time since wetting. Our flammability trials demonstrate that both flame
spread rate and fuel consumption increased with time since wetting but that the
rate of increase varied across these taxa. Only the pines are able to ignite
and support flame spread immediately after wetting (Fig.
\ref{fig:spread_time}). In contrast, \emph{Quercus} exhibited a threshold
response to drying with no ignition prior to 72 hours of drying but once
ignition is possible, it had faster spread rates than the other taxa. The
Cupressaceae, which in our two species have dense litter but low maximum water
absorption had low flammability across moisture levels. This result is
consistent with the low flammability found in an Australian Cupressaceae that
have small litter particles and dense packing citep{Scarff+Westoby-2006,
  Bowman+Haverkamp-2018}. However, the densely packed litter is not a
consistent trait across the family because \citet{Cornwell+Elvira+etal-2015}
found relatively high litter flammability and low litter density for some Asian
Cupressaceae.
